Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-08-22 document/ms.tex
moving likelihood method discussion earlier

@@

Instead, we will reuse the hard work that went into building the catalog.
We must first notice that each entry in a catalog is a representation of the
posterior probability
\begin{eqnarray}\eqlabel{crazypost}
p(\entry_k\,|\,\data_k,\,\interim) &=&
\frac{p(\data_k\,|\,\entry_k)\,p(\entry_k\,|\,\interim)}
     {p(\data_k\,|\,\interim)}
\end{eqnarray}
of the parameters $\entry_k$ conditioned on the observations of that object
$\data_k$.
The notation \interim\ is a reminder that the catalog was produced under a
specific choice of some---probably ``uninformative''---\emph{interim prior}
$p(\entry_k\,|\,\interim)$.
This prior was chosen by the author of the catalog and it is different from
the distribution that we want to infer: $p(\entry_k\,|\,\ratepars)$.

@@

In the inference procedure described above, we only discussed an abstract
occurrence rate density.
To make it more concrete, we'll model the catalog as a draw from the
inhomogeneous Poisson process set by the \emph{observable} rate density
$\obs{\rate}_\ratepars$.
This leads to the previously known result (see \citealt{tabachnik,youdin} for
some of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate
density of exoplanets that you would expect to observe taking into account the
geometric transit probability and any other detection efficiencies.
In practice, we can model the observable rate density as
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency (including transit
probability) at \entry\ and $\rate_\ratepars(\entry)$ is the object that we
want to infer: the \True\ occurrence rate density.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the rate density as, for example, a broken power law or
a histogram.

>>>>

\section{The likelihood method}
\sectlabel{likelihood}

The first ingredient for any probabilistic inference is a likelihood function;
a description of the probability of observing a specific dataset given a set
of model parameters.
In this particular project, the dataset is a catalog of exoplanet measurements
and the model parameters are the values that set the shape and normalization
of the occurrence rate density.
Throughout this \paper, we use the notation $\rate_\ratepars(\entry)$ for the
occurrence rate density \rate---parameterized by the parameters \ratepars---as
a function of the physical parameters \entry\ (orbital period, planetary
radius, \etc).
In this framework, the occurrence rate density can be ``parametric''---for
example, a power law---or a ``non-parametric'' function---such as a histogram
where the bin heights are the parameters \ratepars.

We'll model the catalog as a draw from the inhomogeneous Poisson process set
by the \emph{observable} rate density $\obs{\rate}_\ratepars$.
This leads to the previously known result (see \citealt{tabachnik,youdin} for
some of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate
density of exoplanets that you would expect to observe taking into account the
geometric transit probability and any other detection efficiencies.
In practice, we can model the observable rate density as
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency (including transit
probability) at \entry\ and $\rate_\ratepars(\entry)$ is the object that we
want to infer: the \True\ occurrence rate density.
We haven't yet discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the rate density as, for example, a broken power law or a
histogram.

Finally, for all the results in this \paper, we'll model the rate density as
a piecewise constant step function
\begin{eqnarray}\eqlabel{rate-model}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where the parameters $\ratepar_j$ are the log step heights and the bin edges
are fixed \foreign{a priori}.

In \app{inv-det-eff}, we use the model in \eq{rate-model} and neglect
observational uncertainties, we can derive the analytic maximum likelihood
solution for the step heights.
This result is similar to and just as simple as the
inverse-detection-efficiency method and it is guaranteed to provide a lower
variance estimate of the rate density than the standard procedure.

One major benefit of expressing the problem of occurrence rate inference
probabilistically is that it can now be formally extended to include the
effects of observational uncertainties.

@@

Instead, we will reuse the hard work that went into building the catalog.
We must first notice that each entry in a catalog is a representation of the
posterior probability
\begin{eqnarray}\eqlabel{crazypost}
p(\entry_k\,|\,\data_k,\,\interim) &=&
\frac{p(\data_k\,|\,\entry_k)\,p(\entry_k\,|\,\interim)}
     {p(\data_k\,|\,\interim)}
\end{eqnarray}
of the parameters $\entry_k$ conditioned on the observations of that object
$\data_k$.
The notation \interim\ is a reminder that the catalog was produced under a
specific choice of a---probably ``uninformative''---\emph{interim prior}
$p(\entry_k\,|\,\interim)$.
This prior was chosen by the author of the catalog and it is different from
the likelihood $p(\entry_k\,|\,\ratepars)$ from \eq{poisson-like}.

@@
